---
title: "DataIntergrationChallenge"
author: "Abbi Edwards"
date: "07/06/2021"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(scater)
library(scran)
library(batchelor)
library(bluster)
library(tidyverse)
library(pheatmap)
library(clustree)
library(Cairo)
library(BiocSingular)
library(cowplot)
```

## Loading the data you need


```{r Load}

p.tsne <- readRDS("../Robjects/DataIntergration_p.tsne.rds")
mnn.out <- readRDS("../Robjects/DataIntergration_mnn.out.rds")

clusters.mnn <- readRDS("../Robjects/DataIntergration_clusters.mnn.rds")
linear_rescaled <- readRDS("../Robjects/DataIntergration_linear_rescaled.rds")
rescaled <- readRDS("../Robjects/DataIntergration_rescaled.rds")

```

## Challenge 1

Draw the TSNE for the fastmnn corrected data. Compare it with the uncorrected TSNE, what do you think?

```{r yourCode1}

mp.tsne <- plotTSNE(mnn.out, colour_by="Sample.Name", shape_by = "source_name")
mp.tsne

```

```{r}
mp.tsne + facet_wrap(. ~ mnn.out$source_name)
```

```{r tsne compare}
plot_grid(p.tsne,mp.tsne)
```

```

## Challenge 2

Can you draw the heatmaps for the other samples and decide if you think the correction looks satisfactory?

```{r yourCode, echo=FALSE}
tabE2 <- table(paste("after", clusters.mnn[linear_rescaled$batch=="ETV6-RUNX1_2"]),
               paste("before", colLabels(rescaled[[1]])))
heatE2 <- pheatmap(log10(tabE2 +10), cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "ETV6-RUNX1_2")
tabE3 <- table(paste("after", clusters.mnn[linear_rescaled$batch=="ETV6-RUNX1_3"]),
               paste("before", colLabels(rescaled[[1]])))
heatE3 <- pheatmap(log10(tabE3 +10), cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "ETV6-RUNX1_3")
tabE4 <- table(paste("after", clusters.mnn[linear_rescaled$batch=="ETV6-RUNX1_4"]),
               paste("before", colLabels(rescaled[[1]])))
heatE4 <- pheatmap(log10(tabE4 +10), cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "ETV6-RUNX1_4")
tabP1 <- table(paste("after", clusters.mnn[linear_rescaled$batch=="PBMMC_1"]),
               paste("before", colLabels(rescaled[[1]])))
heatP1 <- pheatmap(log10(tabP1 +10), cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "PBMMC_1")
tabP2 <- table(paste("after", clusters.mnn[linear_rescaled$batch=="PBMMC_2"]),
               paste("before", colLabels(rescaled[[1]])))
heatP2 <- pheatmap(log10(tabP2 +10), cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "PBMMC_2")
tabP3 <- table(paste("after", clusters.mnn[linear_rescaled$batch=="PBMMC_3"]),
               paste("before", colLabels(rescaled[[1]])))
heatP3 <- pheatmap(log10(tabP3 +10), cluster_rows = FALSE, cluster_cols = FALSE,
                   main = "PBMMC_3")
heatE2
heatE3
heatE4
heatP1
heatP2
heatP3



```

