---
title: "CRUK CI Summer School 2020 - introduction to single-cell RNA-seq analysis"
subtitle: 'Dimensionality reduction for analysis'

author: "Stephane Ballereau, Zeynep Kalender Atak, Katarzyna Kania"
#date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
date: July 2020
#bibliography: bibliography.bib
#csl: biomed-central.csl
output:
  html_notebook:
    number_sections: yes
    toc: yes
    toc_float: yes
    fig_caption: yes
    self_contained: true
    fig_width: 6
    fig_height: 4
params:
  projDir: "/ssd/personal/baller01/20200511_FernandesM_ME_crukBiSs2020"
  dirRel: ".."
  inpDirBit: "AnaWiSce/Ana1"
  outDirBit: "AnaWiSce/Ana1"
  cacheBool: FALSE
  bookType: "mk"
  setName: "caron"
  setSuf: "_5hCellPerSpl"
  splSetToGet: "dummy"
  dsiSuf: '_dummy'
---

# Dimensionality reduction for analysis {#dimRedForAnaTop}

<style>
div.blue {background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">
**TODO** add text and fix fig size.
</div>

```{r}
projDir <- params$projDir
dirRel <- params$dirRel
outDirBit <- params$outDirBit
cacheBool <- params$cacheBool
cacheBool <- params$cacheBool
setName <- params$setName
setSuf <- params$setSuf

if(params$bookType == "mk") {
	setName <- "caron"
	setSuf <- "_5hCellPerSpl"
	dirRel <- ".."
}

nbPcToComp <- 50
```

```{r setup_dimRedForAna, include=FALSE, echo=FALSE, dev="CairoPNG"}
require(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE, cache=cacheBool)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(eval = TRUE) 
options(stringsAsFactors = FALSE)
opts_chunk$set(fig.width=7, fig.height=7)
#opts_chunk$set(dev = c("png"))
options(bitmapType='cairo')
knitr::opts_chunk$set(dev="CairoPNG")
set.seed(123) # for reproducibility 
```

```{r libraries_dimRedForAna, message=FALSE, warning=FALSE}
library(ggplot2)
library(scater)
library(scran)
library(dplyr)
library(Cairo)
fontsize <- theme(axis.text=element_text(size=12),
		  axis.title=element_text(size=16))
```

## Load data

We will load the R file keeping the SCE object with the normalised counts for 500 cells per sample.

```{r load_dimRedForAna, cache=FALSE}
# Read object in:
tmpFn <- sprintf("%s/%s/Robjects/%s_sce_nz_postDeconv%s_dimRed.Rds",
		 projDir, outDirBit, setName, setSuf)
print(tmpFn)
if(!file.exists(tmpFn))
{
	knitr::knit_exit()
}
sce <- readRDS(tmpFn)
sce

#any(duplicated(rowData(sce)$ensembl_gene_id))
# some function(s) used below complain about 'strand' already being used in row data,
# so rename that column now:
colnames(rowData(sce))[colnames(rowData(sce)) == "strand"] <- "strandNum"

# also get feature selection outcome
tmpFn <- sprintf("%s/%s/Robjects/%s_sce_nz_postDeconv%s_featSel.Rds",
		 projDir, outDirBit, setName, setSuf)
tmpList <- readRDS(tmpFn)
dec.sce <- tmpList$dec.sce
hvg.index <- tmpList$hvg.index
rm(tmpList)
```

## Denoising expression values using PCA

Aim: use the trend fitted above to identify PCs linked to biology.

Assumption: biology drives most of the variance hence should be captured by the first PCs, while technical noise affects each gene independently, hence is captured by later PCs.

Logic: Compute the sum of the technical component across genes used in the PCA, use it as the amount of variance not related to biology and that we should therefore remove. Later PCs are excluded until the amount of variance they account for matches that corresponding to the technical component. 

```{r comp_denoisePCA, include=TRUE}
var.fit <- metadata(dec.sce)

# remove uninteresting PCs:
#options(BiocSingularParam.default=IrlbaParam())
sce <- denoisePCA(sce, technical=var.fit$trend, assay.type="logcounts")
rm(var.fit)

# check assay names, should see 'PCA':
##assayNames(sce)
# check dimension of the PC table:
##dim(reducedDim(sce, "PCA")) 

sce.pca <- reducedDim(sce, "PCA") # get copy of PCA matrix
tmpCol <- rep("grey", nbPcToComp) # set colours to show selected PCs in green
tmpCol[1:dim(sce.pca)[2]] <- "green"
barplot(attributes(sce.pca)$percentVar[1:nbPcToComp],
        main=sprintf("Scree plot for the %s first PCs", nbPcToComp),
        names.arg=1:nbPcToComp,
        col=tmpCol,
        cex.names = 0.8)

# cumulative proportion of variance explained by selected PCs
round(cumsum(attributes(sce.pca)$percentVar)[1:dim(sce.pca)[2]])

# plot on PC1 and PC2 plane:
plotPCA(sce, colour_by = "Sample.Name")
#require(knitr); knit_exit()
rm(sce.pca, tmpCol)
```

Show cells on plane for PC1 and PC2:

```{r plot_denoisePCA}
plotReducedDim(sce,
	       dimred = "PCA",
	       ncomponents = 3,
	       colour_by = "Sample.Name") +
fontsize
```

## Visualise expression patterns of some HVGs

```{r viz_dimRedForAna}
o <- order(dec.sce$bio, decreasing=TRUE)
chosen.genes.index <- o[1:20]
dec.sce %>%
	data.frame() %>%
	tibble::rownames_to_column("ensembl_gene_id") %>%
	right_join(data.frame(rowData(sce)[chosen.genes.index,]), by="ensembl_gene_id") %>%
	dplyr::select(ensembl_gene_id, Symbol, bio, FDR) %>%
	arrange(-bio)
rm(dec.sce)
```

On PCA plot:

```{r plot_count_pca_HVGtop2}
# make and store PCA plot for top HVG 1:
pca1 <- plotReducedDim(sce,
		       dimred="PCA",
		       colour_by=rowData(sce)[chosen.genes.index[1],"ensembl_gene_id"]) +
fontsize  # + coord_fixed()
# make and store PCA plot for top HVG 2:
pca2 <- plotReducedDim(sce,
		       dimred="PCA",
		       colour_by=rowData(sce)[chosen.genes.index[2],"ensembl_gene_id"]) +
fontsize # + coord_fixed()

pca1
pca2
```

```{r plot_count_pca_HVGtop2_facet, fig.width=12, fig.height=6}
# display plots next to each other:
# multiplot(pca1, pca2, cols=2)

pca1 + facet_wrap(. ~ sce$source_name) + coord_fixed()
pca2 + facet_wrap(. ~ sce$source_name) + coord_fixed()

rm(pca1, pca2)
```

On t-SNE plot:

```{r plot_count_tsne_HVGtop2}
# plot TSNE, accessing counts for the gene of interest with the ID used to name rows in the count matrix:
# make and store TSNE plot for top HVG 1:
tsne1 <- plotTSNE(sce, colour_by=rowData(sce)[chosen.genes.index[1],"ensembl_gene_id"]) + fontsize
# make and store TSNE plot for top HVG 2:
tsne2 <- plotTSNE(sce, colour_by=rowData(sce)[chosen.genes.index[2],"ensembl_gene_id"]) + fontsize

tsne1
tsne2
```

```{r plot_count_tsne_HVGtop2_facet, fig.width=12, fig.height=6}
# display plots next to each other:
#multiplot(tsne1, tsne2, cols=2)

tsne1 + facet_wrap(. ~ sce$source_name)
tsne2 + facet_wrap(. ~ sce$source_name)

# display plots next to each other, splitting each by sample:
#multiplot(tsne1 + facet_grid(. ~ sce$Sample2), tsne2 + facet_grid(. ~ sce$Sample2), cols=2)

rm(tsne1, tsne2)
```

Write R object to file

```{r write_denoised_dimRedForAna}
tmpFn <- sprintf("%s/%s/Robjects/%s_sce_nz_postDeconv%s_denoised.Rds", projDir, outDirBit, setName, setSuf)
print(tmpFn)
saveRDS(sce, file=tmpFn)
```

Genes correlated with PC1:

<!-- http://bioinformatics.age.mpg.de/presentations-tutorials/presentations/modules/single-cell//bioconductor_tutorial.html -->

```{r corel_pc1_dimRedForAna}
sce <- readRDS(tmpFn)
rownames(sce) <- uniquifyFeatureNames(rowData(sce)$ensembl_gene_id,
				      rowData(sce)$Symbol)
pc1 <- reducedDim(sce, "PCA")[,1]
design <- model.matrix(~pc1)
library(limma)
fit <- lmFit(logcounts(sce), design)
fit <- eBayes(fit, trend=TRUE, robust=TRUE)
topTab <- topTable(fit)
head(topTab)
rm(topTab)
```

```{r heatmap_dimRedForAna}
library(pheatmap)
de.genes <- rownames(topTable(fit, coef=2, n=50))
heat.vals <- logcounts(sce)[de.genes,]
heat.vals <- heat.vals - rowMeans(heat.vals)
heat.vals[heat.vals > 2] <- 2
heat.vals[heat.vals < -2] <- -2
p <- pheatmap(heat.vals[,order(pc1)], cluster_cols=FALSE)
#plot(p$gtable)
rm(pc1, design, fit, de.genes, heat.vals)
rm(sce)
```

## Session information

<details>
```{r}
sessionInfo()
```
</details>
